---
title: "Education 250B: Organizational Analysis of Higher Education"
subtitle: "Winter 2021"
author: 
date: 
urlcolor: blue
output: 
  # pdf_document
  # word_document:
  #   toc: no
  #   toc_depth: '3'
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    fig_caption: true
    highlight: tango
    theme: default
    df_print: tibble

---

# Course information

```{r echo=FALSE, message=FALSE}
library(kableExtra)
library(tidyverse)

t <- c("Class website", "Class website password", "Day and Time", "Class Zoom link")
l <- c("https://anyone-can-cook.github.io/orgs_class/", "coalition", "Wednesdays, 8:30AM - 10AM", "https://ucla.zoom.us/j/92479470216?pwd=RnFabUV4bEx2OHBYTm5yZXlNT3I2QT09")

info <- tibble(t, l)

kable(info,col.names = c("", ""), escape = T) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "responsive"), full_width = T, position = "left")
```

# Teaching team

## Instructor

**Ozan Jaquette**  

- Pronouns: he/him/his
- Office: Moore Hall, Room 3038
- Email: [ozanj@ucla.edu](ozanj@ucla.edu)
- Office hours:
  - [Zoom link](https://ucla.zoom.us/j/2545750562)
  - Mondays 3-4pm, and by appointment
  
## Course facilitator

**Demeturie Gogue**  

- Pronouns: he/him/his
- Email: [dtlgogue@g.ucla.edu](dtlgogue@g.ucla.edu)
- Office hours:
  - Email for Zoom link
  - By appointment
  
# Course description

"Organizations" include the schools and colleges we attend, businesses that create the products we use (and change our lifestyle), the bureaucracies that regulate daily life, the non-profit organizations that exert influence on societal policy, etc.  For good or for ill, organizations dramatically affect our lives.  But we are not helpless victims of organizational actions. Sometimes as individuals and as more often as collective actors, we can exert influence over what organizations do

In this class, you will be a student of organizational theory with the goal of understanding organizational behavior. Writ large, organizational theory draws from multiple disciplines (e.g., sociology, psychology, economics, political science, and anthropology) and each discipline approaches organizational behavior from a different perspective and with a different focus.

Most courses on organizational theory draw primarily from psychology and/or sociology.  Organizational theories from psychology tend to focus on more "micro" interactions between people, covering topics such as individual decision-making, implicit bias, motivation, trust, etc.  

Organizational theories from sociology tend to have a more "macro" focus, developing theories to address topics such as: how does the prescribed curriculum of a degree program change (or not change); how do donors control the behavior of departments/universities; how do foundations control the higher education policy agenda and how do other actors influence what problems/solutions are deemed "legitimate" by foundations.

This course focuses primarily on organizational theories from sociology rather than psychology. Why? Because 10 weeks is not enough time to cover both disciplinary perspectives well. And also because I was trained as a sociologist!

The course is organized around theories rather than topics. For example, we don't spend a week studying "organizational change." Rather, several theories we learn throughout the quarter offer different perspectives on organizational change. 

## What is theory?

What is "theory"? A lot of ink has been spilled over this question. Here's a simple, working definition: A theory is an explanation that can be applied to more than one situation.  You will develop a deep understanding of several of the core sociological theories of organizational behavior (e.g., "old" institutionalism, new institutionalism, resource dependence theory).

## Course goals

1. Understand the theory
    - Gain a deep understanding of the core ideas and concepts from sociological theories of organizational behavior
2. Understand the strengths and weaknesses of each theory
    - This goal requires that you fully engage with the theory - it is not acceptable to reject a theory/article because "I didn't like it" or "It does not focus on X"
    - I want your criticisms of the theory to be deep/substantive criticisms. Deep criticisms only come from deep understanding.
    - I also want you to think about what is useful and insightful about the theory.
3. Identify connections and areas of disagreement across theories
    - Identify common concepts utilized by different theories
    - Where do two theories disagree with one another?
    - Theories help you simplify clutter by telling you what to focus on. How does the "focus" of one theory (i.e., which relationships/factors are deemed important) differ from the focus of another theory?
4. Apply theory to practice
    - Practice applying theory to current/previous personal experiences as a student or at your job
    - Practice applying theory to contemporary issues happening at UCLA and in the broader field of education
5. Begin to apply theory to research

# Course structure

The course will generally focus on a particular theoretical perspective each week. Class will consist of weekly synchronous lectures on Wednesdays from 8:30AM to 10AM and an asynchronous component to be completed before the live class.

1. **Asynchronous lectures materials and reading**. Weekly lecture slides and associated pre-recorded lecture videos (hopefully 60 minutes or less) will be posted on the [course website](https://anyone-can-cook.github.io/orgs_class/), along with the required and encouraged readings for the week. Students are expected to work through the lecture material and complete the readings in advance of our weekly synchronous class meeting. Lectures are designed to provide context to help you understand the required readings better and/or will discuss main ideas from the required readings.
2. **Synchronous class session**. We will have one synchronous class meeting per week on Wednesdays from 8:30AM to 10AM. What we do during this 90-minute period will vary week to week, but generally there will be a student-led activity/discussion around the topic of the week followed by time for Q&A about the reading and lecture.

## Using cases to compare theories

Over the coming weeks, we will introduce three "cases" that we will use to compare one theory to another. These cases will be the focus of essay questions in your two written assessments during the quarter.

Of particular interest, most of the dominant theoretical perspectives on the sociology of organizations were written by white men and often assuming heterosexual white male actors. How well do these theories perform when the protagonist does not fit this description? Can we get a better understanding of our cases by applying critical race-forward or gender-forward theoretical perspectives? Can very different theoretical perspectives by integrated to complement their respective strengths? These are the sorts of questions we will ponder as we apply theory to our three cases.

The three cases are associated with these three pieces of scholarship:

- Hevel, M. S. (2017). _Toward a history of LGBT student organizations' struggle-for-recognition lawsuits: The exemplar of GLSA v. Gohn._
- Cole, E. R. (2020). _The campus color line: College presidents and the struggle for Black freedom_. Princeton, New Jersey: Princeton University Press.
- Taylor, K.-Y. (2017). _How we get free: Black feminism and the Combahee River Collective_. Chicago, Illinois: Haymarket Books.

# Communication and course website

## Course website

All course-related material can be found on the [course website](https://anyone-can-cook.github.io/orgs_class/). Pre-recorded lecture videos, lecture slides, and readings will be posted on the class website under the associated section. Weekly course topics can also be found on the class website.

The course website is encrypted to avoid making copyright protected article/chapter PDFs public. The password to decrypt the website is: **coalition**.

## Questions/discussion about course related content

We will vote on our mode of communication during the first class.

## Communication with instructor and facilitator

If you have a personal question or issue, you can email the instructor or course facilitator directly. Additionally, we are available for office hours or by appointment if there is anything you would like to discuss with us in private.

# Classroom environment

We all have a responsibility to ensure that every member of the class feels valued and safe. Be mindful that our words and body language affects others in ways we might not fully understand. We have a responsibility to express our ideas in a way that doesn't make disparaging generalizations and doesn't make people feel excluded. As an instructor, I am responsible for setting an example through my own conduct.

Learning organizational theory can be hard, especially when reading dense articles from sociology or org theory journals. I struggled with much of the material when I was first introduced to it, and sometimes I still struggle. It is important for all of us to help create an environment where students feel safe asking questions about what ideas/concepts they struggled with. This is how we learn.


# Course readings

This course is reading intensive. There will generally be three or four -- often dense  -- required readings per week. Typically, this will amount to about 80 - 100 pages of required reading each week.


All required and optional readings for each week will be listed on the [course website](https://anyone-can-cook.github.io/orgs_class/).

**Required books** (any edition fine)

- Brint, S. G., & Karabel, J. (1989). _The diverted dream: community colleges and the promise of educational opportunity in America, 1900-1985_. New York: Oxford University Press. 
- Pfeffer, J., & Salancik, G. R. (1978). _The external control of organizations: A resource dependence perspective_. New York: Harper & Row.
- Taylor, K. Y. (2017). _How we get free: Black feminism and the Combahee River Collective_. Chicago, Illinois: Haymarket Books.

**Optional books**

- Birnbaum, R. (1991). _How colleges work: The cybernetics of academic organization and leadership_. San Francisco: Jossey-Bass.
- Scott, W. R., & Davis, G. F. (2006). _Organizations and organizing: Rational, natural, and open systems perspectives_. Upper Saddle River, New Jersey: Pearson, Prentice Hall.

# Assignments and grading

**Class participation and attendance** (15%)

- Because this course is a seminar, class discussion, participation, and attendance are vitally important.
- Show up to class on time (me too!)
- Read all required readings prior to class and be prepared to ask questions about areas of confusion.
- Be engaged in class discussion. This includes participating, as well as listening to and helping your classmates.

**Student-led activity for theory application** (15%)


- Each week, a group of (usually) two students will be responsible for leading a 30-45 minute in-class activity that allows the class to apply the theory from the week’s readings to real-world scenarios. 
- The goal of these activities is to move beyond simply learning theory to envisioning how these frameworks emerge in higher education as well as other organizational structures. 
- Class activities should complement the course lectures and readings as well as engage other organizational theories that we will be learning throughout the quarter.
- [Guidelines for leading class discussion](https://docs.google.com/document/d/1VCCR2ouyhxK4bFvmKcHpbOQGSS-ou9jruilpMoVYZMk/edit?usp=sharing)" document (subject to change at discretion of instructors)

**Mid-quarter three-page essay** (10%)

- The mid-quarter essay prompt will be distributed after class on week 3 or week 4 and it will be due two weeks later.
- The purpose of this assignment is to help you prepare for the take-home final in two ways:
    - Gain familiarity with the sort of essay questions I will ask you to answer for the take-home final
    - Become familiar with my expectations for writing quality (e.g., organization of ideas, prose, etc.)

**Take-home final exam** (60%)

- The take-home final will be distributed after class on week 10 and it will be due one week later (no readings will be assigned that week)
- The take-home final will assess the following sorts of things:
    - Demonstrate understanding of core concepts about the theories we learn
    - Describe important limitations/blind spots of theories we learn
    - Comparing similarities and differences across theories
    - Ability to apply different theories to develop insights about a particular issue in higher education
- We will devote class time in the second half of the quarter (including substantial class time in week 10) to preparing for the take-home final.
- Preparation for the take-home final will prepare you for the DAP exam.

## Grading policy

**Attendance**

- Each unexcused absence results in a loss of 2% from your overall grade for the course. Two or more unexcused absence will result in a failing grade for the course. An excused absence is a professional opportunity or religious holy day that you discuss with me beforehand or a medical or family emergency. Excused absences will not result in a loss of attendance points. However, you will be responsible for all material covered in that class.
- Please show up to class on time.

**Guidelines for written work**

- Use APA style. Written assignments should have 1 inch margins, double spacing, size 11 or 12 font.
- Late assignments will lose 10% for the first day late and an additional 10% for each week late. Assignments that are late due to medical or family emergencies will not receive a deduction in points. Contact me as soon as possible. If the reason is legitimate, I will be understanding and accommodating.
	
**Learning during a pandemic**

With the ongoing spread of the COVID-19 pandemic, we understand that right now is a challenging and precarious time for everybody. Many of us may be experiencing added stress or responsibilities that make learning and completing classwork difficult. If you are having trouble keeping up with the class, please reach out to the teaching team and we will help work out a plan with you. We want to be accommodating to everyone's unique situation and we hope to avoid any deductions to your grade.


## Grading scale

```{r echo=FALSE, message=FALSE}
lg <- c("A", "A-", "B+", "B", "B-", "C+", "C", "C-")
nv <- c("93+%", "90<93%", "87<90%", "83<87%", "80<83%", "77<80%", "73<77%", "70<73%")

grade <- tibble(lg, nv)

kable(grade,col.names = c("Letter Grade", "Percentage"), escape = T) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F, position = "left")
```

# Course policies

## Zoom guidelines

All synchronous class sessions will be held online, via Zoom. Below, we have outlined some general guidelines about Zoom learning. As we continue learning together, we can add to and change the below list. I'm open to your feedback and your experiences as we continue to learn how to learn via Zoom.  

- **Video**: We will not require students to turn on their video during synchronous lectures. We encourage students to turn on their video only if they feel comfortable doing so -- particularly during small group breakout rooms.  
- **Audio**: We ask students to mute their microphones when they are not speaking. We encourage the use of earphones or headphones if you are in a space with background noise.  
- **Zoom outage**: In the unlikely event that Zoom is down, the instructors will email the class with instruction for completing the class section in an asynchronous manner. Therefore, if Zoom is not functioning properly during the class period, be sure to check your email often.  
- **Internet connectivity**: We understand that having access to a stable internet connection and/or electronic equipment is a privilege. With that in mind, we want to provide a space where everyone has the resources they need to do well in the class. If you have any issues with your internet connection and/or don't have access to electronic equipment, please reach out to the instructors.   

## Academic accomodations

**Center for Accessible Education**  

Students needing academic accommodations based on a disability should contact the Center for Accessible Education (CAE). When possible, students should contact the CAE within the first two weeks of the term as reasonable notice is needed to coordinate accommodations. For more information visit https://www.cae.ucla.edu/.

Located in A255 Murphy Hall: (310) 825-1501, TDD (310) 206-6083; http://www.cae.ucla.edu/

-	Due to COVID-19, the CAE office is closed for in-person meetings
-	CAE counselor, resources, and services are still available via email / virtual appointment
-	Stay up-to-date with CAE newsletters & announcements at https://www.cae.ucla.edu/announcements-events/student

## Academic integrity

__UCLA policy__  

  - UCLA is a community of scholars. In this community, all members including faculty, staff and students alike are responsible for maintaining standards of academic honesty. As a student and member of the University community, you are here to get an education and are, therefore, expected to demonstrate integrity in your academic endeavors. You are evaluated on your own merits. Cheating, plagiarism, collaborative work, multiple submissions without the permission of the professor, or other kinds of academic dishonesty are considered unacceptable behavior and will result in formal disciplinary proceedings.  
  
__This class__  

  - Given that much of the grade for this course depends on the take-home final, take steps to avoid accidental plagiarism. Make sure you cite your sources and use quotation marks when you are explicitly copying text from another author. It can be easy to accidentally plagiarize when you are summarizing the work of another scholar. So when you are taking notes on a paper, make sure to put everything in your own words or put copied text in quotation marks.

# Campus resources


## Counseling and Psychological Services (CAPS)

As a student you may experience a range of issues that can cause barriers to learning, such as strained relationships, increased anxiety, alcohol/drug problems, depression, difficulty concentrating and/or lack of motivation. These mental health concerns or stressful events may lead to diminished academic performance or reduce a student's ability to participate in daily activities. UC offers services to assist you with addressing these and other concerns you may be experiencing. If you or someone you know are suffering from any of the aforementioned conditions, consider utilizing the confidential mental health services available on campus.

Students in distress may speak directly with a counselor 24/7 at (310) 825-0768, or may call 911; located in Wooden Center West; https://www.caps.ucla.edu

-	CAPS is open and has transitioned to Telehealth services ONLY
-	Open Mon – Thurs: 8am-6pm and Fri: 8am-5pm
-	As always, 24/7 crisis support is always available by phone at (310) 825-0768


## Discrimination

UCLA is committed to maintaining a campus community that provides the stronget possible support for the intellectual and personal growth of all its members- students, faculty, and staff. Acts intended to create a hostile climate are unacceptable.  

  - To file an online incident report, visit: https://equity.ucla.edu/report-an-incident/  


## LGBTQ resource center

The LGBTQ resource center provides a range of education and advocacy services supporting intersectional identity development. It fosters unity; wellness; and an open, safe, inclusive environment for lesbian, gay, bisexual, intersex, transgender, queer, asexual, questioning, and same-gender-loving students, their families, and the entire campus community. Find it in the Student Activities Center, or via email lgbt\@lgbt.ucla.edu.  

  - Visit their website for more information: https://www.lgbt.ucla.edu/ and virtual upcoming events

## International students

The Dashew Center provides a range of programs to promote cross-cultural learning, language improvement, and cultural adjustment. Their programs include trips in the LA area, performances, and on-campus events and workshops.  

  -	Due to COVID-19, the Dashew Center has transitioned its operations to a remote setting
  - Visit their website for more information: https://www.internationalcenter.ucla.edu/  
  - For COVID updates, visit https://www.internationalcenter.ucla.edu/covid-19-updates

## UCLA Undocumented Student Program

This program provides a safe space for undergraduate and graduate undocument students. USP supports the UndocuBruin community through personalized services and resources, programs, and workshops.  

  - Visit their website for more information: https://www.usp.ucla.edu/  
  - You can reach USP at usp@saonet.ucla.edu

## Student legal services

UCLA Student Legal Services provides a range of legal support to all registered and enrolled UCLA students. Some of their services include:  

  - Landlord/Tenant Relations (Including challenges during COVID)
  - Accident and Injury Problems
  - Domestic Violence and Harassment
  - Divorces and Other Family Law Matter  

Due to COVID, Student legal Services is closed to walk-ins.  

  - All services are by appointment only
  - For more information visit their website: http://www.studentlegal.ucla.edu/index.php  
  

## Students with Dependents

UCLA Students with Dependents provides support to UCLA studens who are parents, guardians, and caregivers. Some of their services include:  

- Information, referrals, and support to navigate UCLA (childcare, family housing, financial aid)
- Access to information about resources within the larger community
- On-site application and verification for CalFresh (food stamps) & MediCal and assistance with Cal Works/GAIN
- A quiet study space
- Family friendly graduation celebration in June  

For more information visit their website: https://www.swd.ucla.edu/

## Campus maps 

**Lactation Rooms** 

- [Map to lactation rooms on campus](https://ucla.app.box.com/v/2019-lactation-map)  

**Gender Inclusive restrooms**  

- [Map to gender inclusive restrooms](https://www.lgbt.ucla.edu/Portals/38/Documents/GIRR%20Map%2007-2019%20Compressed.pdf)  

**Campus accessibility**  

- [Campus accessibility map](http://map.ucla.edu/downloads/pdf/Access_08_21_15.pdf)


## Title IX Resources

Title IX prohibits gender discrimination, including sexual harassment, domestic and dating violence, sexual assault, and stalking.  If you have experienced sexual harassment or sexual violence, there are a variety of resources to assist you.  

- __CONFIDENTIAL RESOURCES__:You can receive confidential support and advocacy at the CARE Advocacy Office for Sexual and Gender-Based Violence, A233 Murphy Hall, CAREadvocate\@careprogram.ucla.edu, (310) 206-2465. Counseling and Psychological Services (CAPS) also provides confidential counseling to all students and can be reached 24/7 at (310) 825-0768.  

- __NON-CONFIDENTIAL RESOURCES__: You can also report sexual violence or sexual harassment directly to the University's Title IX Coordinator, 2255 Murphy Hall, titleix\@conet.ucla.edu, (310) 206-3417. Reports to law enforcement can be made to UCPD at (310) 825-1491. These offices may be required to pursue an official investigation.

*Faculty and TAs are required under the UC Policy on Sexual Violence and Sexual Harassment to inform the Title IX Coordinator should they become aware that you or any other student has experienced sexual violence or sexual harassment.*  
